---
title: "Rapa Nui Mataa Morphometric Analyses"
author: "Carl P. Lipo, Rene Horneman, Terry L. Hunt, Vincent Bonhomme"
date: "3 September 2014"
output: 
  pdf_document:
    keep_tex: true
---


```{r, error=FALSE, message=FALSE,   echo=FALSE, cache=FALSE, warning=FALSE}
require(knitr)
opts_chunk$set(fig.align='left') 
require(devtools)
install_github("vbonhomme/Momocs")
require(Momocs)

```

# Introduction

Rapa Nui (Easter Island, Chile) is a tiny island located in a remote corner of Eastern Polynesia, more than 2000 km from the nearest inhabited body of land. The island was first colonized by Polynesians who sailed from central East Polynesia in voyaging canoes during the 13th century AD (Hunt and Lipo 2008; Wilmshurst et al. 2011). Depsite the island's diminuitive size, remoteness and limited natural resources, the prehistoric Rapanui are well-known for the construction of nearly 1000 mult-ton statues known as *moai* (REFS). The contrast between the spectacular nature of the archaeological record and the sparse enviornment of the island has led some researchers to believe that prehistoric populations  grew in numbers until resource use exceeded the carrying capacity of the island, resulting in catastropic demographic collapse.  New research, however, is now challenging this notion with new evidence that the Rapanui flourished on the island until AD 1722 when Europeans brought diseases and other social disruptions (Hunt 200X; Hunt and Lipo 20XX, 2011; Mulrooney 20XX). 


One of the claims related to the "collapse" scenario is that  prehistoric Rapa Nui populations experienced intense warfare during late prehistory when resources became increasingly scarce (Bahn and Flenley 1992; Diamond 1995, 2005; Flenley and Bahn 2002). In addition to oral traditions recorded in the 20th century (Metraux 1940; Routledge 1917), one line of evidence used to make this assumption is the presence of *mata’a*, flaked obsidian stemmed tools. *Mata’a* are a class of hafted flaked obsidian artifacts that are found commonly on Rapa Nui. As stemmed obsidian tools, their form is similar to artifacts found on other islands such as New Britain, Papua New Guinea (e.g., Torrence et al. 2008).

*Mata'a* are one of the most numerous shaped artifact classes found on the surface of Rapa Nui.  these artifacts have been noted since the earliest European visitors described the island.  For example, Johann Georg Adam Forster, the son of a German naturalist who accompanied Captain Cook on his second voyage to the Pacific, reported that on Rapa Nui individuals “had lances or spears made of thin ill-shaped sticks, and pointed with a sharp triangular piece of black glassy lava” (von Saher 1990:35).

A great diversity of shape has remained one particularly puzzling aspect of mata`a. Mata`a shapes are highly inconsistent and vary from rounded to subangluar to angular to complex.  The earliest researchers assigned mata`a shape variation to what they conceived as ethnographic categories based on Rapanui words (Routledge 1919).  Later attempts to construct systematic classifications have also focused on identifying types based on characterizations of overall shape.  None of these classification efforts produced useful categories.

Mulloy (1961:151), for example, argued that “no significant clustering or correlations could be extracted.... the material represents a continuous range of variation without objective natural order, and that the only classification possible must involve the subjective selection of ideal types from infinite series of possibilities, and the arbitrary reference of intermediate for to one or another of these.”  Mulloy concluded manufacturing procedures dictated shapes and differences in overall shape of mata’a were best explained by chance.

Overall shape is rarely a useful dimension for problem-oriented classification.  The forms of objects are limited by technological constraints of the material, performance aspects that depend upon the range of environments in which the object is used, and simple idiosyncratic variability related to the manufacturer and the process of production.  In the case of mata’a much of the variability in the overall blade shape can be explained by the idiosyncratic stages of manufacture (Bolt et al 2006) plus differences structured by functional variation related to the range and kinds of activities for which the tool was primarily used: cutting, scraping, or some combination (as shown in use-wear studies, see Church and Rigney 1994; Church and Ellis 1996).  



Previous stylistic analysis and evaluation of usewear for *mata'a* contradict claims of weapons and instead point to their use as either some kind of symbolic implement or their use in cultivation activities (Bollt et al. 2006; Lipo et al. 2010). [[ NEED MORE DETAIL HERE]]

In this analysis, we use shape morphometric analysis to examine *mata’a* shape variability and to look for areas of shape that are constrained versus those that are free to vary. Our results conclude that mata’a were only functionally constrained in terms of the haft. These results continue to support the alternative hypotheses that these artifacts were not used as weapons.   

As a group, *mata'a* vary greatly in size and shape, but average 6-10 cm in width and length. They formed from a hard hammer flakes with the haft formed from one of the lateral margins of the flake and the blade from the remaining distal and opposite lateral margin. The haft and shoulder is formed from unifacial flakes and is generally lenticular in form. Overall, the blade shape is dominated by the shape of the parent flake though some additional shaping can be evident. Often, large areas of cortex still cover much of one face. much of one face. 

We assume that due to performance the functional aspects of any tool result in shape variability that is more constrained than the non-functional or stylistic attributes.  Based on this notion, we hypothesise that :
   * If *mata’a* are weapons, the distal end of the artifact will be constrained. However, if *mata’a* are not weapons, other areas of the tool will show greater constraint  consistent with alternate functions.
   * If *mata’a* are weapons, the distal end of the artifact will show a tendency towards a pointed spear-like shape that will penetrate either enemies or prey. If *mata’a* are not weapons, there will be no such constrictive tendency at the distal end of the tool.
   * If there was inter-tribal warfare, *mata’a* from distinct areas may show stylistic traits of distinct groups. If natives were not divided into warring groups, distinct stylistic  traits may or may not be apparent. 

# Data


```{r, data-analysis, echo=FALSE}

data.path <- "../Data/RapaNuiMataaData.txt" 
# We read the whole file
PAST <- read.table(data.path, header = TRUE, sep="\t")
last.meta <- which(colnames(PAST)=="X1") - 1
fac <- PAST[, 1:last.meta]
# xy will contain coordinates only
xy <- as.matrix(PAST[, -c(1:last.meta)])
# a short loop to reorder thing and store them in a list
coo <- list()
for (i in 1:nrow(xy)){ 
  coo[[i]] <- cbind(xy[i, seq(1, ncol(xy), 2)], xy[i, seq(2, ncol(xy), 2)])}
# we renames the components of the list (ie the shapes)
names(coo) <- fac[, 1]

# now we create the Out object (formerly Coo,
# but Coo became a super-class now, to handle outlines, open outlines, and
# landmarks)
RapaNui <- Out(coo, fac=fac)

## some descriptive information
numberOfMataa <- nrow(xy)
#maximum width
xMaxLength<-0
for (i in 1:nrow(xy)){ xMaxLength[i]<-(max(xy[i, seq(2, ncol(xy), 2)])-min(xy[i, seq(2, ncol(xy), 2)]) ) }

#maximum width
xMaxWidth<-0
for (i in 1:nrow(xy)){ xMaxWidth[i]<-(max(xy[i, seq(1, ncol(xy), 2)])-min(xy[i, seq(1, ncol(xy), 2)]) ) }

# some descriptions of the numbers of mataa
 
englertMataa <- length(which(fac[3] == "Englert"))
bishopMataa <- length(which(fac[3]=="Bishop"))
```


```{r kable, results="asis"}

byCollection <- table(fac$Collection,fac$Site)

bySource <- table(fac$Source,fac$Site)

kable(byCollection, digits=0)
kable(bySource,digits=0)
```



In order to test these hypotheses, we used morphometric outline analysis. With this technique exact outlines of artifacts can be overlaid and compared making variations between artifacts readily apparent. Our assemblage consisted of planview photographs of (N='r numberOfMataa') *mata’a* from 2 museum collections. The first museum collection consisted of XXX *mata'a* housed at the P. Sebastian Englert Museum on Rapa Nui.  This collection is composed of XX *mata'a* collected from XX locations on the island and XXX *mata'a* for which provience is known only to the level of the island itself. [[More detail on the locations?]]

The second collection of *mata'a* is composed of XXX objects housed at the Bishop Museum, Honolulu, Hawai'i. 

of t of XXX mata'a These *mata'a* wer taken by Lipo on Rapa Nui from 2007 through 2009.  25 to 33 artifacts were randomly selected from each of (4) sites. 

# Analyses

Using TPSdig software (Rohlf 2014), each mata’a was measured using 200 sets of X-Y coordinates from the perimeter of the artifact. Radial measurements were also taken every 2 degrees around the circumference. 

We transferred the data to PAST 3.01 (Hammer et al. 2001)  software and X-Y coordinate and radii analyses were done. The X-Y coordinates were used to create overlying images showing shape variability. The radii analysis were used to compute the standard deviation of each radius. The standard deviations were then used to graph out areas of constraint and/or variability in each group.

Each set of analysis were performed separately for each site, and collectively for the entire assemblage.  

```{r, pca-prep, echo=FALSE, cache=TRUE, warning=FALSE}

panel(RapaNui, fac = "Site")
panel(RapaNui, fac = "Island")
panel(RapaNui, fac="Source", reorder="Source")

# Lets begin by centering, scaling them 
stack(RapaNui)

RapaF <- eFourier(RapaNui, 12)
RapaP <- fgProcrustes(RapaNui)
RapaFP <- eFourier(RapaP, 12, norm=FALSE)


```


## By Source
```{r, rapaf-plot,fig.width=4.5,fig.height=4.5}

plot(PCA(RapaF), "Source")

```


```{r, rapafp-plot,results='asis'}
plot(PCA(RapaFP), "Source") # Procrustes aligned (normalization of the outlines)
```

## By Site
```{r, rapafs-plot, results='asis'}
plot(PCA(RapaF), "Site") # regular EFT with normalized coefficients
```

```{r, rapafps-plot, results='asis'}
plot(PCA(RapaFP), "Site") # Procrustes aligned (normalization of the outlines)
```


Some embedded plots.

```{r, rapafp-p, results='asis'}
plot(PCA(RapaFP), "Site") # Procrustes aligned (normalization of the outlines)
```

# References Cited

Bollt, R., Clark, J.E., Fisher, P.R., Yoshida, H.K., 2006. An experiment in the replication and classification of Easter Island mata`a. Rapa Nui Journal 20:125-133.
Church, F. Ellis, J.G., 1996.  A use-wear analysis of obsidian tools from an Ana Kionga. Rapa Nui Journal 10, 81-88.
Church, F., Rigney, J., 1994. A microwear analysis of tools from Site 10-241, Easter Island—an inland processing site. Rapa Nui Journal 8, 101-105.
Diamond, J., 2005. Collapse: How societies choose to fail or succeed, Viking, New York.
Eerkens, J.W., Lipo, C.P. 2005. Cultural transmission, copying errors, and the generation of variation in material culture and the archaeological record.  Journal of Anthropological Archaeology 24(4):316-334.
Flenley, J., Bahn, P., 2003. The Enigmas of Easter Island: Island on the Edge. Oxford University Press. Oxford.
Hunt, T.L. and Lipo, C.P., 2007. Chronology, deforestation, and “collapse:” Evidence vs. faith in Rapa Nui prehistory.  Rapa Nui Journal 21(2): 85-97.
Hunt, T.L. and Lipo, C.P. 2009. Revisiting Rapa Nui (Easter Island) “Ecocide.”  Pacific Science 
63(4):601-616.
Kirch, P.V., 1984. The Evolution of the Polynesian Chiefdoms, Cambridge University Press, Cambridge.
Kirch, P.V., 2000. On the Road of The Winds:  An Archaeological History of the Pacific Islands before European Contact, University of California Press, Berkeley.
Lipo, C. P., 2001.  Science, Style and the Study of Community Structure: An Example from the Central Mississippi River Valley. British Archaeological Reports, International Series, no. 918, Oxford.
Lipo, C.P., Hunt, T.L., 2005. Mapping prehistoric statue roads on Easter Island, Antiquity 79, 158-168.
Lipo, C., Madsen, M., Dunnell, R., Hunt, T., 1997. Population Structure, Cultural Transmission, and Frequency Seriation, Journal of Anthropological Archaeology 16, 33.
Mulloy, W., 1961. Ceremonial Center of Vinapu, in: Heyerdahl, T., Ferdon, E.N.J. (Eds.), Archaeology of Easter Island, The School of American Research and the Museum of New Mexico, Santa Fe, pp. 93-161.
Neiman, F. 1995. Stylistic variation in evolutionary perspective: Inferences from decorative diversity and interassemblage distance in Illinois Woodland ceramic assemblages.  American Antiquity 60(1): 7-36.
O’Brien, Michael J., Lyman R. L. 2003 Cladistics and Archaeology. University of Utah Press, Salt Lake City. 
Routledge, K., 1919. The Mystery of Easter Island, Sifton, Praed & Co., London.
von Saher, H., 1990. Some details from the journal of Captain Bouman on the discovery of Easter Island. Rapa Nui Journal 6, 34-39.

