---
title: "Rapa Nui Mataa Morphometric Analyses"
author: "Carl P. Lipo, Rene Horneman, Terry L. Hunt, Vincent Bonhomme"
date: "7 September 2014"
output:
  word_document:
    reference_docx: reference.docx
  pdf_document:
    keep_tex: yes
bibliography: bibliography.bib
email: carl.lipo@csulb.edu
csl: antiquity.csl
affiliations:
  '1': California State University Long Beach
  '2': University of Oregon
  '3': French Institute of Pondicherry
---

```{r, setup, error=FALSE, message=FALSE,   echo=FALSE, cache=TRUE, warning=FALSE}

#suppressPackageStartupMessages(library(rCharts, quietly=TRUE))
library(knitr)
library(httr)
library(Rmisc)
library(lattice)
library(plyr)
library(plotrix)
#set_config( config( ssl.verifypeer = 0L ) )
opts_chunk$set(fig.align='left') 
options(repos = c(CRAN = "http://cran.rstudio.com"))
set_config( config( ssl.verifypeer = 0L ) )
library(devtools)
if("Momocs" %in% rownames(installed.packages()) == FALSE) { install_github("vbonhomme/Momocs")}
library(Momocs)
```


```{r, data-analysis, echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE}
# some constants we use in the file
data.path <- "../Data/xyfinaldatawithids.txt" 
scalingFactor = 75 ## this is the # of pixels per cm. 
xCenter = 12 # x coordinate of centroid (if its fixed, otherwise can be aligned based on average centroid, etc.)
yCenter = 11 # y coordinate of centroid

# We read the whole file
allPAST <- read.table(data.path, header = TRUE, sep="\t")
PAST <-allPAST[which(allPAST$Island=="Rapa_Nui"),]
nbPAST <-PAST[which(allPAST$Island=="New_Britain"),]

last.meta <- which(colnames(PAST)=="X1") - 1
fac <- PAST[, 1:last.meta]
allfac<-allPAST[,1:last.meta]

# xy will contain coordinates only
xy <- as.matrix(PAST[, -c(1:last.meta)])
allxy<- as.matrix(allPAST[,-c(1:last.meta)])
scaledXY <- xy/scalingFactor
allscaledXY <-allxy/scalingFactor

# a short loop to reorder thing and store them in a list
coo <- list()
allcoo <-list()
for (i in 1:nrow(scaledXY)){ 
  coo[[i]] <- cbind(scaledXY[i, seq(1, ncol(scaledXY), 2)], scaledXY[i, seq(2, ncol(scaledXY), 2)])
  }

for (i in 1:nrow(allscaledXY)){ 
  allcoo[[i]] <- cbind(allscaledXY[i, seq(1, ncol(allscaledXY), 2)], allscaledXY[i, seq(2, ncol(allscaledXY), 2)])
  }
# we renames the components of the list (ie the shapes)
names(coo) <- fac[, 1]
#names(allcoo) <-allfac[,1]

# now we create the Out object (formerly Coo,
# but Coo is now a super-class now in order to handle outlines, open outlines, and
# landmarks)
RapaNui <- Out(coo, fac=fac)

## include the non Rapa Nui examples
allMataa <-Out(allcoo, fac=allfac)

## Descriptive information for the paper
numberOfMataa <- nrow(scaledXY)

# Create vectors of the lengths and widths for descr. stats. 
## note that in the images 1 cm = 75 pixels
lengths <- 0
widths <- 0
for (i in 1:nrow(scaledXY)){ 
  lengths[i] <- (max(scaledXY[i, seq(2, ncol(scaledXY), 2)])-min(scaledXY[i, seq(2, ncol(scaledXY), 2)]))
  widths[i] <- (max(scaledXY[i, seq(1, ncol(scaledXY), 2)])-min(scaledXY[i, seq(1, ncol(scaledXY), 2)]))
}

# descriptive values
meanLength <-mean(lengths)
sdLength <-sd(lengths)
meanWidth <- mean(widths)
sdWidth <-sd(widths)

# Determine maximum width and length
xMaxLength<-0
xMaxWidth<-0
for (i in 1:nrow(scaledXY)){ 
  xMaxLength[i]<-(max(scaledXY[i, seq(2, ncol(scaledXY), 2)])-min(scaledXY[i, seq(2, ncol(scaledXY), 2)]) )
  xMaxWidth[i]<-(max(scaledXY[i, seq(1, ncol(scaledXY), 2)])-min(scaledXY[i, seq(1, ncol(scaledXY), 2)]) )
  }

# Create some descriptions of the numbers of mataa in different categories
englertMataa <- length(which(fac[3] == "Englert"))
bishopMataa <- length(which(fac[3]=="Bishop"))
unknownMataa <- length(which(fac[3]=="Unknown"))
numberOfSites <-length(fac[4])-1


# the dmp function
dmp <- function(q, xlab="", ylab="", title="", th0=pi/2,
                cols, palette=col.summer, leg, ...){
  op <- par(mar=c(5.1, 0, 4.1, 0))
  q <- q/max(q)
  nr <- nrow(q)
  nc <- ncol(q)
  #th <- seq(0, 2*pi, length=nc+1)[-(nc+1)] + th0
  th <- seq(0, 2*pi, length=nc) + th0
  xi <- t(apply(q, 1, function(r) r*cos(th)))
  yi <- t(apply(q, 1, function(r) r*sin(th)))
  if (missing(cols)) cols <- palette(nr)
  plot(NA, asp=1, xlim=c(-1, 1), ylim=c(-1, 1),
       xlab=xlab, ylab=ylab, main=title,
       xaxs="i", las=1, ann=FALSE, axes=FALSE, frame=FALSE)#, ...)
  lines(cos(th), sin(th), col="grey80")
  for (i in seq(0, 1, 0.25)){
    lines(i*cos(th), i*sin(th), col="grey80", lty=2)}
  th.grid <- seq(0, 2*pi, length=9)
  segments(0, 0, cos(th.grid), sin(th.grid), col=c("grey80"), lty=2)
  for (i in 1:nr) lines(xi[i, ], yi[i, ], col=cols[i])
  if (missing(leg)) leg <- rownames(q)
  legend("topright", lwd=1, col=cols, legend=leg, cex=3/4, bty="n")}

## plot the confidence intervals
ci.plot <- function(x){
  #x <- lapply(x$coo, function(x) apply(x, 1, ed, pt2=c(-xCenter, -yCenter)))
  x <- lapply(x$coo,coo.interpolate,360)
  x <- matrix(unlist(x), nrow=length(x), byrow=TRUE)
  xCI <- apply(x, 2, CI,ci = 0.99)
  ## now add some and substract to differentiate between the 3 lines.. this is just arbitrary spacing.
  l<-xCI[1,]-1
  m<-xCI[2,]
  u<-xCI[3,]+1
  xxCI=rbind(l,m,u)
  dmp(xxCI)
  }

quartiles.plot<-function(x){
  #x$coo <- lapply(coo.trans$coo, x$coo, - 625, - 500)
  #x <- lapply(x$coo, function(x) apply(x, 1, ed, pt2=c(-xCenter, -yCenter)))
  x <-lapply(x$coo,coo.interpolate,360)
  x <- matrix(unlist(x), nrow=length(x), byrow=TRUE)
  quartilesCI <- apply(x, 2, quantile, probs=seq(0, 1, 0.25))
  dmp(quartilesCI)
  }

```

# Introduction

Rapa Nui (Easter Island, Chile) is a tiny island located in a remote corner of Eastern Polynesia, more than 2000 km from the nearest inhabited body of land (Figure 1). The island was first colonized by Polynesians who sailed from central East Polynesia in voyaging canoes during the 13th century AD [@Hunt:2006aa; @Wilmshurst:2011aa]. Depsite the island's diminuitive size, remoteness, and limited natural resources, the archaeological record of Rapa Nui is well-known for its nearly 1000 multi-ton statues known as *moai* that once sat atop massive stone platforms known as *ahu* [@Hunt:2011aa]. 

![Figure 1.  Location of Rapa Nui (Easter Island, Chile) in Polynesia.](../Figures/Figure1.pdf)

Figure 1. Location of Rapa Nui in Polynesia.

The dramatic prehistoric investment in monumental architecture stand in marked contrast to the Rapa Nui's environment and historically observed population levels. Even at the first point of European contact, the tiny island was largely devoid of trees and population sizes were just about 3000 individuals [@Hunt:2011aa, p. XXX]. While earlier researchers [e.g., @Heyerdahl:1965aa; @Heyerdahl:1989aa] believed the depleted and depauperate state of the island was due to conflict between Polynesians and elite from South America, more recent reseachers have interepreted the contrast between the spectacular nature of the archaeological record and the sparse enviornment of the island as the outcome of a prehistoric environmental catastrophe [@Bahn:1992aa; @Flenley:2003aa].  These researchers argue that based largely on oral traditions, that prehistoric populations grew in numbers until resource use exceeded the carrying capacity and the island underwent catastropic demographic collapse. This account has been popularlized as the "collapse" scenario [*sensu* @Diamond:1995aa; -@Diamond:2005aa].

New research, however, has challenged this notion with empirical evidence generated from the archaeological record that the Rapanui flourished on the island until AD 1722 when Europeans brought diseases and other social disruptions [@Hunt:2007aa; @Hunt:2007ab; @Hunt:2009aa; @Hunt:2009ac; @Hunt:2011aa; @Hunt:2011ab; @Lipo:2009ab; @Mulrooney:2009aa; @Mulrooney:2012aa; @Rainbird:2002aa]. Investigations on Rapa Nui's settelement patterns demonstrates that the island's inhabitants lived in a dispersed pattern in a low density fashion [@Hunt:2011aa; @Morrison:2012aa]. In addition, studies show that subsistence was largely based on extensive but marginally productive lithic mulch gardens to boost the nutrient-poor soil to a level that sustained sweet potato cultivation [@Bork:2004aa; @Ladefoged:2005aa; @Ladefoged:2010aa; @Ladefoged:2013aa; @Mieth:2010aa; @Stevenson:2002aa; @Stevenson:2006aa]. Finally, demise of the once extensive palm tree forest appears to have had little to do with statue construction or changes in carry capacity [@Hunt:2011aa; @Lipo:2013aa].

One of the claims that persists that is thought to support the "collapse" scenario is the idea that prehistoric Rapa Nui populations experienced intense warfare during late prehistory when resources became increasingly scarce [@Bahn:1992aa; @Diamond:1995aa, @Diamond:2005aa; @Flenley:2003aa]. Oral traditions are known that attribute the toppling of stone statues to intertribal prehistorc warfare [@Bahn:1992aa]. But the existence of fallen statues alone does not necessarily imply warfare since other natural explanations are more likely [@Edwards:1996aa]. Indeed, the existing evidence points to the toppling of statues as as series of post-contact historic events rather than prehistory [@Hunt:2011aa]. Most significantly, examples of defensive structures are entirely lacking in the island's archaeological record [@Hunt:2011aa; @Lipo:2014aa].  Overall, much of the evidence for prehistoric warfare among the inhabitants of Rapa Nui comes from oral traditions recorded in the 20th century [e.g., @Routledge:1919aa]. The oral traditions, however, have an unknown relation to prehistory. Metraux [-@Metraux:1940:aa, p. XXX], for example, argues that most of the traditions are likely recent and thus likely do not reflect prehistoric events. Given the unknown origins of oral traditions, we must rely upon direct archaeological evidence for warfare. 

The one example of empirical evidence used to support arguments about prehistoric warfare on Rapa Nui is the presence of *mata’a*, flaked obsidian stemmed tools. *Mata’a* are a class of hafted flaked obsidian artifacts that are found commonly on Rapa Nui. As relatively simple stemmed obsidian tools with wide blades, their form is similar to artifacts known as *mata* found on other Polynesian islands such the basalt artifacts found on New Zealand, Pitcairn and the Chatham Islands [@Balfour:1917aa; @Metraux:1957aa, p. 232; @Skinner:1958aa] as well as New Britain, Papua New Guinea [e.g., @Araho:1997aa;  @Specht:1988aa; @Torrence:2009aa; @Torrence:2009ab; @Torrence:2013aa].

In the current analysis, we seek to explore whether there exists variability in the shape of *mata'a* that sheds provides information about the fucntional environment in which these artifacts interacted.  Using a large image database of 'r numberOfMataa' *mata'a* from Rapa Nu, we conduct quantitative morphometric analyses to further investigate whether specific tool classes might be identifiable in the range of shapes in which these artifacts are found. Morphometric analyses enable on to explore shape as a continuous property of objects rather than requiring us treat shape as nominal categories. In this way we can use principal components analyses to see of particular kinds of shapes map to particular locations, environments or source material. In addition, we can examine the relative patterns *mata’a* shape variability and to look for areas of shape that are constrained versus those that were more free to vary. Overall, our results conclude that mata’a were only functionally constrained in terms of the haft and had signficant variation on the distal end and blade.  These results continue to support the alternative hypotheses that these artifacts were not used as weapons. The degree of similarity, however, of the haft portion of *mata'a* and the low degree of constraint in the blade poses an intriguing puzzle: we have yet to identify the role(s) that these objects played in Rapa Nui subsistence and settlement.  

# Approach

*Mata'a* have been noted since the earliest European visitors described the island.  Members of Cook's expedition to the island commented that the islanders “had lances or spears made of thin ill-shaped sticks, and pointed with a sharp triangular piece of black glassy lava” [@Saher:1990aa, p. 35]. *Mata'a* are often assumed to be "spears" largely because of their resemblance to European varieties rather than any direct observation of their use.  Scars noted by early European observers are also believed to have been inflicted by *mata'a* though there is no clear evidence that their use was lethal.  For example, in his voyage to Rapa Nui in 1770, Captain Don Felipe González [@Haedo:1908aa, p. 99] remarked that "they [Rapanui] possess no arms, and although in some we observed sundry wounds on the body, which we thought to have been inflicted by cutting instruments of iron or steel, we found that they proceeded from stones, which are their only [weapons of] defence and offence, and as most of these are sharp edged they produce the injury referred to."

Even if we had direct observations of these objects being used in "spear-like" fashion, the unavoidable tendency for these European observers to interpret what they saw through their own preconceptions requires us examine the physical evidence available on mata'a. In this way we can learn not only the range of interactions that the objects had with the environment but also determine if there is variablity in their use through time or over space.

On the surface landscape of Rapa Nui, *mata'a* are one of the most numerous shaped artifact classes. Overall, *mata'a* vary greatly in size and shape, but average 6-10 cm in width and length. Technologically, they are formed from unifacial flakes derived trough hard hammer percussion on obsidian cores quarried from one of the island's obsidian sources.  Most of the shaping of the *mata'a* occurs during the creation of a stem that presumably serves as a haft. The stem is formed from one of the lateral margins of the original flake where blade constitutes the remaining distal and opposite lateral margins. *Mata'a* stems and shoulders are formed by unifacial flaking and are generally lenticular in cross section. Overall, the blade shape is dominated by the shape of the parent flake though some shaping through secondary flaking is sometimes evident. Often, large areas of cortex still cover much of one face.

In exploring the the way in which *mata'a* forms vary, researchers have noted that there is a great diversity of shape [@Mulloy:1961aa]. *Mata`a* shapes are highly inconsistent and vary from rounded to subangluar to angular to complex.  Early researchers assigned mata`a shape variation to what they conceived as ethnographic categories based on Rapanui words [i.e., @Routledge:1919aa].  Later attempts to construct systematic classifications have also focused on identifying types based on characterizations of overall shape.  None of these classification efforts produced useful categories.

Mulloy [-@Mulloy:1961aa, p. 151], for example, argued that “no significant clustering or correlations could be extracted.... the material represents a continuous range of variation without objective natural order, and that the only classification possible must involve the subjective selection of ideal types from infinite series of possibilities, and the arbitrary reference of intermediate for to one or another of these.”  Mulloy concluded manufacturing procedures dictated shapes and differences in overall shape of mata’a were best explained by chance.
 
The overall shape of an object is rarely a useful dimension for problem-oriented classification [@Dunnell:1986aa]. The forms of objects are limited by technological constraints of the material, performance aspects that depend upon the range of environments in which the object is used, and simple idiosyncratic variability related to the manufacturer and the process of production.  In the case of *mata’a* much of the variability in the overall blade shape can be explained by the contingent results involved in the stages of manufacture [@Bollt:2006aa]. The difference in shapes, therefore, may have structured  functional variation related to the range and kinds of activities for which the tool was primarily used.  Studies of use-wear found on *mata'a* point to the tool being used primarily for scraping and cutting or some combination  [@Church:1994aa; @Church:1996aa]. 

A recent study of *mata'a* shape using stylistic classes and deterministic frequency seriation as a means for examining how class frequencies changed over space and through time showed remarkably continuous change [@Lipo:2010aa]. The seriation results suggest that the source of variability in *mata'a* form is largely being inherited through the social learning of manufacturing techniques between indivdiuals.  The evidence also indicates that variability in the form of *mata'a* is not related to how the *mata'a* performed in its use environment(s).   Overall, our growing understanding of *mata'a* variability continues to support their form being related to ceremonial or cultivation activities and not as weapons invovled in warfare [@Bollt:2006aa; @Lipo:2010aa].

In our analysis here, we focus on *mata'a* variability in the blade portion of the mata'a relative to the stem.  We assume that as hafted objects the point at the center of the stem where it meets the blade can be held constant for comparisons of shape.  We then assume that due to performance the functional aspects of the tool will result in shape variability that is more constrained than the non-functional or stylistic attributes [@Lipo:2012aa]. The constraints are the result of natural selection that serves to sort shape variability in proportion to the benefits/drawbacks to performance. Based on this notion, we hypothesise that:

   - If *mata’a* are weapons, the distal end of the artifact will be constrained. However, if *mata’a* are not weapons, other areas of the tool will show greater constraint  consistent with alternate functions.
   
   - If *mata’a* are weapons, the distal end of the artifact will show a tendency towards a pointed spear-like shape that will penetrate either enemies or prey. If *mata’a* are not weapons, there will be no such constrictive tendency at the distal end of the tool.
   
   - If there was inter-tribal warfare, *mata’a* from distinct areas may show stylistic traits of distinct groups. If natives were not divided into warring groups, distinct stylistic  traits may or may not be apparent. 

# Methods and Data

In order to test these hypotheses, we used morphometric outline analysis.  Morphometrics is the quantitative analysis of form in terms of shape and size [@Bookstein:1982aa; @Bookstein:1985aa; @Bookstein:1997aa; @Cardillo:2010aa; @Kendall:1989aa; @Rohlf:1990aa]. It has advantages over traditional studies of shape that treat shape as a nominal character (e.g., "triangular", "square", "round"). Even classifications that break shape into a series of dimensionally constructed classes reduce variability into modal categories. Morphometrics avoid the problem of nominal shape by analyzing the form of objects as a series of metric measurements that characterize the relative positions of series of landmarks or comprise the outline.  Analysies of form variability can be conducted in two and three dimensions [@Kendall:1989aa]. With techniques available for standardizing scale and rotation, morphometric measurements directly compare outlines of artifacts and generate data on the variations between artifacts.  Consequently, one major feature of morphometrics is its ready ability to statistically test hypotheses about the factors that affect shape.  

Measurements for morphometrics can be generated in a number of ways.  With roots in biology, the earliest form of morphometrics focused on identifying the location specific landmarks [e.g., @Thompson:1917aa]. A landmark approach requires defining features of interest that are to be examined as to how they relate to each other.  In the case of artifacts such as *mata'a* there are few consistent landmarks to hold constant other than perhaps the distal and proximal end. One can also one can conduct an analysis of what is known as "semi-landmarks," a fixed number of regularly positioned points around the outline of an object [@Bookstein:1997aa; @Gunz:2013aa]. Both approaches to measuring shape make use of the relative positions between all points [@Bookstein:1991aa; -@Bookstein:1997aa].
  
In our morphometric analyses we make use of Momocs (http://CRAN.R-project.org/package=Momocs), an R package (R Core Team 2014) developed by Bonhomme [-@Bonhomme:2012aa; @Bonhomme:2014ab]. Momocs builds upon techniques developed by Claude [-@Claude:2008aa] and reviewed by Bowman [-@Bowman:2009aa]. Bonhomme incorporated functions from Claude's work into an integrated framework and a standalone R package. The package's vignette *A Graphical Introduction to Momocs and Outline Analysis Using R* [@Bonhomme:2012aa]  provides an extensive description of the functions of the package.

Table 1:  *Mata'a* included in analyses by collection.
```{r kable1, results="asis", message=FALSE, echo=FALSE, warning=FALSE}
byCollection <- table(fac$Collection,fac$Site)
bySource <- table(fac$Source,fac$Site)
knitr:::kable(byCollection, digits=0)
```

Table 2:  *Mata'a* included in analyses by location.
```{r kable2, results="asis", message=FALSE, echo=FALSE, warning=FALSE}
knitr:::kable(bySource,digits=0)
```

For our analyses of Rapa Nui *mata'a*, our assemblage consisted of planview photographs of (N='r numberOfMataa') artifacts from two museum collections. Outlines of the studied *mata'a* are shown in Figure 3).  The first museum collection consisted of `r englertMataa`  *mata'a* housed at the P. Sebastian Englert Museum on Rapa Nui.  This collection is composed of photograps of *mata'a* collected from `r numberOfSites` locations on the island as well as `r unknownMataa ` *mata'a* for which provience is known only to the level of the island itself (Figure 2). 

The second collection of *mata'a* is composed of `r bishopMataa` objects housed at the Bishop Museum, Honolulu, Hawai'i. These *mata'a* consist of examples  purchased from the island by a private collector in 1920, collections made by Kenneth P. Emory in 1929-1931 and various gifts to the museum [@Mulrooney:2014aa, p. 5-6]. Mulrooney and colleagues took photos of these *mata'a* during their study of obsidian sourcing via pXRF [@Mulrooney:2014aa]. Their findings demonstrate that the majority of *mata'a* were made from obsidian obtained at the Orito source, though some examples are made from obsidian from the Rano Kao and offshore Motu Iti sources. Due to the manner in which the Bishop Museum material was originally collected, we can only attribute the source of these *mata'a* to Rapa Nui and and not a specific location on the island.   Mulrooney and colleagues, however, kindly provided source identifications for each *mata'a* based on the results of their study. In this way, we are able to use these *mata'a* examples to examine potential shape variability that might be due to the obsidian source. This shape variability could be potentially caused by systematic material differences or by the differential use of *mata'a* that are dervied from different locations. 

![Figure 2.  Locations of *mata'a* collections from Rapa Nui, Chile.](../Figures/Figure2.pdf)

Figure 2. Locations of *mata'a* collections from Rapa Nui, Chile.

```{r family_picture, message=FALSE, echo=FALSE, warning=FALSE, fig.width=4.5,fig.height=4.5}
Momocs:::panel(RapaNui, fac = "Collection")
#Momocs:::panel(RapaNui, fac = "Site")
#Momocs:::panel(RapaNui, fac = "Island")
#Momocs:::panel(RapaNui, fac="Source", reorder="Source")
```

Figure 3.  *Mata'a* included in the current analyses.

For the purposes of our analyses and following the approach taken in Lipo et al. [-@Lipo:2010aa], we assumed that *mata'a* are the hafted portions of compound tools that are otherwise incompletely preserved in archaeological deposits.  In this context and based on evidence of usewear on the distal edges [@Church:1994aa; @Church:1996aa], we assume that the overall shape of *mata'a* shape is a functional element [sensu @Dunnell:1978aa], the portion of the artifact that interacts with the environment.  Consequently, our interest is on those aspects of shape that potentially affect function and thereby come under natural selection.  The task of explaining variability in shape consists of identifying selective pressures that affect the performance of shape and to determine whether their magnitude is sufficiently great to impact fitness.  The greater the selective pressures on performance, the more constraint we would expect on those aspects of shape.  If the effect on function and performance is sufficiently small, then other forces such as technological (i.e., material source, manufacturing steps, etc.) or stylistic (stochastic or neutral) ones may have played a role in fixing the shapes of *mata'a*, as well as when and where they occur in the archaeological record. In these cases, we would expect to see a greater range of variability. It is possible, however, that not all *mata'a* instances were used in the same way.  If *mata'a* shapes is influcenced by more than one function, either contemporaneously or over time, then the selective context will differ and thus the “cause” of *mata'a* shape should vary. This situation should create modal patterns of mata'a where shape variability forms statistically distinguishablke groups. 

# Data

In our analyses, we aligned scaled photos of *mata'a* at the point where the midpoint of the stem meets the blade. We oriented each image with the centerline of the haft. These steps helped insure that the shapes were directly comparable. We converted the images to binary to isolate the artifact from the background and then used TPSdig software [@Rohlf:2014aa] to create outlines of each *mata’a*.  TPSDig was particularly useful as it provides a means for automatically tracing outlines with a fixed number of points.  In the creation of outlines, we identified 200 sets of X-Y coordinates located on equidistant points along the perimeter of the artifact (Figure 4). The set of coordinates for each mata'a were aggregated into a single file using PAST [@Hammer:2001aa]. Finally, we standardized the outlines for differences in sizes by dividing the cooridinate positions by the total length of each outline. 

![Figure 4.  Measurement process used to generate semilandmark data for each *mata'a* image.](../Figures/Figure4-StepsInMeasurement.pdf)

Figure 4. Measurement process used to generate semilandmark data for each *mata'a* image.

```{r length_metrics, message=FALSE, echo=FALSE, warning=FALSE, fig.width=4.5,fig.height=4.5}
par(mfrow=c(1, 2))
hist(lengths, main="Length (cm)") 
hist(widths, main="Width (cm)") 
par(mfrow=c(1, 1))
```

Figure 5. *Mata'a* length and width.

While the range of variability in shape shown in Figure 4 is substantial, simple metrics of length and width (Figure 5) suggest that there is just a single distribution of these objects without clear-cut modes in size. Comparisons of length and width, however, are fairly crude descriptions of shape.  A more direct means of evaluating shape variability is accomplished by superimposing  *mata'a* outlines (Figure 6). This process required selecting a standard reference point for all objects from which measurements would be based.  We selected our reference points, referred to here as “centroids,” based on the points from which we believe variability will be meaningfully constrained (or not).  In this case of *mata'a*, we chose a centroid at the center of the haft where it intersects the blade.    

```{r stack, message=FALSE, echo=FALSE, warning=FALSE, fig.width=4.5,fig.height=4.5}
# Lets stack them scaling them 
stack(RapaNui)
```

Figure 6. Superimposed *mata'a* outlines from Rapa Nui.  For comparison, all *mata'a* are aligned at the center point of the haft where it meets the blade. 

Once we identified the centroid, we calculated the distance from the centroids to the perimeter in one-degree intervals for the 360-degree perimeter.  One-degree increments  provide sufficient detail about shape at a scale that characterized overall shape variability with enough detail to capture attributes regarding the haft shape and distal blade shape outline.   

While our measured outlines are composed of 200 points, Momocs interpolates between points to locate distances from centroids at even intervals.In addition, since all measurements are based on georeferenced coordinates, planimetric measure (such as width or length) can be calculated.  Additional image analysis techniques to isolate object outlines point to the strong potential for automation of the measurement process, greatly increasing the ability to characterize large assemblages. With large numbers of measures of radial distances made relative to the *mata'a* centroids, we then calculate a statistical summary for each angle to assess variability in relative dimensions (Figure 7).  

```{r width_metrics, message=FALSE, echo=FALSE, warning=FALSE, fig.width=4.5,fig.height=4.5}
# we interpolate before equianular sampling
#RapaNui$coo <- lapply(RapaNui$coo, coo.interpolate, 360)
#RapaNuirr.int <- coo.samplerr(RapaNui, 360) # and plot this one
par(mfrow=c(1,1))
#ci.plot(RapaNuirr.int)
ci.plot(RapaNui)
#quartiles.plot(RapaNuirr.int)
quartiles.plot(RapaNui)
par(mfrow=c(2, 1))
```

Figure 7.  Variability in *mata’a* shape shown with mean and 95% confidence intervals. Note that the 95% confidence intervals are shown with exaggerated differences between the values to areas with greater variance versus those with more constrained shape.

```{r polar_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.width=4.5,fig.height=4.5}
#polar.plot(u,angles,pos=NULL,label.pos=NULL, start=0,clockwise=TRUE,rp.type="l",lty=3, line.col=3) 
#par(new=T)
#polar.plot(m,angles,pos=NULL,label.pos=NULL, start=0,clockwise=TRUE,rp.type="l",line.col=4) 
#par(new=T)
#polar.plot(l,angles,pos=NULL,label.pos=NULL, start=0,clockwise=TRUE,rp.type="l",lty=3, line.col=5) 
#par(new=F)
```

#Morphometric Analyses

```{r, pca-prep, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
RapaF <- eFourier(RapaNui, 12, norm=TRUE)  #norm makes sure they are scaled
## RapaP <- fgProcrustes(RapaNui) # Not sure we need this as all of our objects are aligned
```

Some text here about the various ways in which morphometrics can be done.

##Elliptic Fourier Analysis

Fourier based analyses are powerful tools for the study of shape variability [@Claude:2008aa]. Fourier approaches decompose shape into a periodic function that is the sum of simple trigonometric functions such as sine and cosine. These simple functions are frequencies that are integer multiples, i.e. are harmonics, of one another. Lower harmonics provide approximation for the coarse-scale trends in the original periodic function while the high-frequency harmonics fit its fine-scale variations (Figure 8). In morphometrics, Fourier treats closed outlines as periodic functions. If you start somewhere on the outline and follow it, you will pass again and again by the same starting point and thus periodic functions can describe this outline. These functions can uses a variety of descriptive data for  the outline: the distance of any point on the outline to the centroid of the shape, the variation of the tangent angle for any point, or the (x/y) position on the plane [@Rohlf:1984aa]. For an outline shape, a periodic function is obtained and can be decomposed (and thus described) by Fourier series. 

```{r, ellipic-fourier, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE,  fig.width=4.5,fig.height=4.5}
Ptolemy(RapaNui[1], nb.h = 5, t = seq(0, 2*pi, pi/2), legend = TRUE)
```

Figure 8. For all positive integers, the sum of a cosine curve and a sine curve defines an ellipse in the plane. Elliptic Fourier analysis is based on an harmonic sum of such ellipses. Five harmonics are here shown at four locations on the original outline of a *mata'a*. As the number of harmonics is increased the reconstruction better approximates the original shape outline.

Fourier series, however, work on continuous functions. Since in practice shape is measured on a finite number of discrete points on a plane (in our case, x/y coordinates), a discrete equivalent to Fourier series is used in morphometrics. A given number of points called pseudolandmarks have to be sampled along the outline before computing shape analysis. All Fourier decomposition then result in an harmonic sum of trigonometric functions associated with harmonic coeffi-
cients. They are (usually) normalized to remove homothetic, translational or rotational differences between shapes. Two or four coefficients, depending on the approach used, are obtained for each calculated harmonic and can then be considered as quantitative variables. Nyquist frequency impedes us from having more harmonics than half the number of points fitted, which is thus the upper limit. The geometrical information contained in the outlines are thus quantified and can be analysed with classical multivariate tools.

#Harmonic Power

To conduct Fourier analysis, we must estimate the number of necessary harmonics after examining the spectrum of harmonic Fourier power. The power is proportional to the harmonic amplitude and can be considered as a measure of shape information. As the rank of a harmonic increases, the power decreases and adds less and less information. We can evaluate the mininum number of harmonics required to best approximate the shape. In the case of the *mata'a* and using x/y position for points on the outline as the data set, 12 harmonics provide a good reconstruction of the overall shape (Figures 9 and 10).

```{r, hqual-plot,echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE, fig.width=4.5,fig.height=4.5}
hqual(RapaNui, method = "eFourier", id = 16, title="*Mata'a* reconstructions with harmonics", harm.range = 1:36,  palette = col.sari, plot.method = "panel")
```

Figure 9. *Mata'a* reconstructed from different numbers of harmonics. Twelve harmonics provide a satisfactory reconstruction.

```{r, hpow-plot,echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE, fig.width=4.5,fig.height=4.5}
hpow(RapaNui, method = "efourier", title="Fourier coefficients power spectrum")
```

Figure 10.  Cumulated harmonic Fourier power calculated from Rapa Nui *mata'a*. The 12 first harmonics gather nearly 100% of the harmonic power. Maxima, minimaRapa and medians are also plotted.

Based on the elliptical Fourier characterizations of the *mata'a* we can now examine the shapes to determine if there are clusters of shape classes that might distinguish different functional sub-groups from each other.  Figure 11 presents the  position of mata'a shapes on a factorial map with shapes reconstructed from first two principal compoonent axes.  Based on these data we do not have discrete shape groups: *mata'a* are highly variable in outline shape and there are continuous intermediate shapes between all variants. This finding demonstrates that the similarity of the proximal haft end of the *mata'a* and that the distal end is free to vary.  These results fail to indicate any subgroup that might have been specifically built as lethal weapons and supports the notion that mata'a have no particular function for which blade shape affects performance. Given observations of usewear on the blade edges (e.g.,@Church:1998aa; @Church:1994aa; @Church@1996aa; @Stevenson:2008aa, p. 107), it is likely that this means that *mata'a* simply must have an edge sufficient for cutting and scraping. 

```{r, rapaf-all-plot,echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE, fig.width=4.5,fig.height=4.5}
# By Site
plot(PCA(RapaF)) 
```

Figure 11.  First two principal components (PC1 and PC2 are on the x- and y-axis, respectively) for the Rapa Nui *mata'a*.

We can also examine the *mata'a* to see if there are systematic differences between the locations from which *mata'a* are found or between the obsidian sources used to make the artifacts. In our analyses, we have examples of *mata'a* from 'r numberOfSites' sites (Table 4).

Table 4:  Rapa Nui *mata'a* included in analyses by site.
```{r kable4, results="asis", message=FALSE, echo=FALSE, warning=FALSE}
bySite <- table(fac$Collection,fac$Site)
knitr:::kable(bySite, digits=0)
```

Figure 12 presents the distribution of sets of *mata'a* from multiple locations across the island. The graph includes the confidence ellipses for the 'r numberOfSites' sites on the grid representing the morphological space based on the first two principal components.  The overlap of the groups indicates that the shapes from each of the sites are cannot be distinguished.  The same conclusion can be drawn from an analysis of the shape variability relative to obsidian source (Figure 13).  Overall, there appears to be no evidence that *mata'a* blade shape was constrained due to functional performance.   

```{r, rapaf-bysite-plot, results='asis', echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE, fig.width=4.5,fig.height=4.5}
## By Source
plot(PCA(RapaF), "Site") # regular EFT with normalized coefficients
```

Figure 12.  First two principal components of *mata'a* grouped by site location. 

```{r, rapaf-bymaterial-plot, results='asis', echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE, fig.width=4.5,fig.height=4.5}
## By Source
plot(PCA(RapaF), "Source") # regular EFT with normalized coefficients
```

Figure 13.  First two principal components of *mata'a* grouped by site location. 

```{r, pca-prep-all, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
allF <- eFourier(allMataa, 12, norm=TRUE)  ## note that this is scaled
```

#Comparison with stemmed tools from other Pacific Islands

The *mata'a* of Rapa Nui share some similarity with stone tools found on other islands across the Pacific.  On New Britain in Melanesia, for example, Torrence [-@Torrence@2009aa,  -@Torrence@2009ab, -@Torrence@2013aa[] has described obsidian stemmed tools that share much in common with the shape of *mata'a.* Torrence (see also @Kononenko:2012aa) argues that these tools may have been used for range of activities including tatooing and ritual scarification.  Pitcairn Island is remote Eastern Polynesian island that   is located ca. 1900 km to the west of Rapa Nui. The island was inhabited by historically related Polynesian populations, though abandoned during prehistory.  An additional but limited comparison can be made for Pitcairn Island, where a few stemmed lithic tools have been found [@Heyerdahl:1965aa]. 

As a comparison for our study, we generated outlines of examples stemmed lithic tools from published images and drawings of these islands using the same procedure for Rapa Nui. While the sample sizes of the non-Rapa Nui assemblages are small, when we compare the shapes of *mata'a* with those other objects, we find that the Pitcairn Island stemmed artifacts have overall shapes that are quite distinct. While we cannot rule out the possibility that the Pitcairn examples are a few extremely long and pointed shapes that happen to have been collected from a much wider array of variability, these shapes are certainly more consistent with hafted tools for hunting or weapons.  

The New Britain artifacts, on the other hand,  wide array of features that are more like the ones from Rapa Nui. Based on this comparison, it is certainly conceivable that one of Rapa Nui *mata'a* reflect the same kinds of uses that are thought to characterize the New Britain tools. Tatooing is known from Rapa Nui through ethnohistoric observation [@Huish:1839aa, p. 77; @Metraux:1940aa; @Thomson:1891aa, p. 22] and as markings on the prehistoric *moai* [@Lee:1992aa]. It would not be surprising that at least some of the mata'a objects were used in tatooing and scarification practices, though their numbers and widespread distribution suggests that they were  likely not all used in this fashion. 

```{r, allfp-plot,echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE,  fig.width=4.5, fig.height=4.5 } 
 plot(PCA(allF), "Island") #amp.shp=1.5)
```

Figure 14:  Factorial maps depicting the two principal compoents (PC1 and PC2 are the x- and y-axis, respectively) of morphological variation for stemmed lithic shaped objects from Rapa Nui, New Britain and Pitcairn Island. The shapes are reconstructed from the factorial map using the first two component axes.

#Conclusion

While the notion that Rapa Nui prehistory consists of a tale of collapse and self-destruction remains popular, the evidence to support this claim is non-existent. In addition to lack of defensive structures and skeletal evidence of lethal violence [@Hunt:2011aa], the so-called "weapons of mass destruction" [@Keegan:1993aa] take a careful look at the shape of *mata'a* we simply do not see evidence that these class of artifacts represent lethal weapons [see also @Ingersoll:2013aa]. There appears to be no performance requirements that would influence the blade shape. Other than having an edges, mata'a are no more lethal than throwing any other kind of rock.  Indeed, as documented in post-contact Rapa Nui, rock throwing from high points is the primary way in which native Rapanui fought off the instrusion of Europeans, [e.g., Roggeveen's 1722 visit, @Eyzaguirre:2004aa].

That *mata'a* had more than one function is not surprising since it is essential to resist the notion that any object is imbibed with an inherent function [@Dunnell:1978aa]. Instead, we measure function on the empirical variability for any assemblage of objects [@Dunnell:1976aa]. In the case of *mata'a* the wear patterns and distribution in rock mulch suggest that at least some of these objects were employed in cultivation. We also cannot rule out that they were used in tattooing and scarification practices.  The latter function certainly is consistent with observations of healed scars made by Spanish visitors in 1770 [@Corney:1908aa].

It is unfortunate that in the case of Rapa Nui, the myth of the island continues to persist despite the lack of evidence. Tradition has long trumped empirical inquiry when it comes to the understanding the functional role that *mata'a* played within the subsistence and social dimensions of prehistoric Rapa Nui. In the case of Rapa Nui, getting the correct answer is far from a trivial academic exercise. The island's prehistory is often used as an exemplar of ignoring the impacts humans made on the environment and thus is driving policy decisions. United Kingdom Prime Minister Margaret Thatcher, for example, famously used Rapa Nui as a warning in a 1989 presentation to the General Assembly of United Nations  [e.g., @Thatcher:2014aa]. Similarly,  *Mata'a* have been used as examples of mass effect "weapons" in a study of terrorist tactics [@Rasmussen:2010aa].  Given the contemporary importance that Rapa Nui has in guiding our concerns for our future, we owe it to ourselves to make certain that we fully understand the prehistory of the island and that our understanding is based on well documented evidence. 

#References Cited

