---
title: "Rapa Nui Mataa Morphometric Analyses"
author: "Carl P. Lipo, Rene Horneman, Terry L. Hunt, Vincent Bonhomme"
date: "3 September 2014"
output: pdf_document
---

```
```{r, echo=FALSE}
require(knitr)
opts_chunk$set(fig.align='left') 
require(devtools)
install_github("vbonhomme/Momocs")
require(Momocs)

```

# Introduction

Rapa Nui (Easter Island, Chile) is a tiny island located in a remote corner of Eastern Polynesia, more than 2000 km from the nearest inhabited body of land. The island was first colonized by Polynesians who sailed from central East Polynesia in voyaging canoes during the 13th century AD (Hunt and Lipo 2008; Wilmshurst et al. 2011). Depsite the island's diminuitive size, remoteness and limited natural resources, the prehistoric Rapanui are well-known for the construction of nearly 1000 mult-ton statues known as *moai* (REFS). The contrast between the spectacular nature of the archaeological record and the sparse enviornment of the island has led some researchers to believe that prehistoric populations  grew in numbers until resource use exceeded the carrying capacity of the island, resulting in catastropic demographic collapse.  New research, however, is now challenging this notion with new evidence that the Rapanui flourished on the island until AD 1722 when Europeans brought diseases and other social disruptions (Hunt 200X; Hunt and Lipo 20XX, 2011; Mulrooney 20XX). 

![Easter Island, Chile (Rapa Nui)](/Volumes/Macintosh HD/users/clipo/Documents/mataaMorphometrics/Figures/Figure1.pdf).

One of the claims related to the "collapse" scenario is that  prehistoric Rapa Nui populations experienced intense warfare during late prehistory when resources became increasingly scarce (Bahn and Flenley 1992; Diamond 1995, 2005; Flenley and Bahn 2002). In addition to oral traditions recorded in the 20th century (Metraux 1940; Routledge 1917), one line of evidence used to make this assumption is the presence of *mata’a*, flaked obsidian stemmed tools. *Mata’a* are a class of hafted flaked obsidian artifacts that are found commonly on Rapa Nui. As stemmed obsidian tools, their form is similar to artifacts found on other islands such as New Britain, Papua New Guinea (e.g., Torrence et al. 2008).

Previous stylistic analysis and evaluation of usewear for *mata'a* contradict claims of weapons and instead point to their use as either some kind of symbolic implement or their use in cultivation activities (Bollt et al. 2006; Lipo et al. 2010). [[ NEED MORE DETAIL HERE]]

In this analysis, we use shape morphometric analysis to examine mata’a shape variability and to look for areas of shape that are constrained versus those that are free to vary. Our results conclude that mata’a were only functionally constrained in terms of the haft. These results continue to support the alternative hypotheses that these artifacts were not used as weapons.   

As a group, *mata'a* vary greatly in size and shape, but average 6-10 cm in width and length. They formed from a hard hammer flakes with the haft formed from one of the lateral margins of the flake and the blade from the remaining distal and opposite lateral margin. The haft and shoulder is formed from unifacial flakes and is generally lenticular in form. Overall, the blade shape is dominated by the shape of the parent flake though some additional shaping can be evident. Often, large areas of cortex still cover much of one face. much of one face. 

We assume that due to performance the functional aspects of any tool result in shape variability that is more constrained than the non-functional or stylistic attributes.  Based on this notion, we hypothesise that :
   * If *mata’a* are weapons, the distal end of the artifact will be constrained. However, if *mata’a* are not weapons, other areas of the tool will show greater constraint  consistent with alternate functions.
   * If *mata’a* are weapons, the distal end of the artifact will show a tendency towards a pointed spear-like shape that will penetrate either enemies or prey. If *mata’a* are not weapons, there will be no such constrictive tendency at the distal end of the tool.
   * If there was inter-tribal warfare, *mata’a* from distinct areas may show stylistic traits of distinct groups. If natives were not divided into warring groups, distinct stylistic  traits may or may not be apparent. 

# Mata'a Analyses

```
```{r}

data.path <- "/Volumes/Macintosh HD/users/clipo/mataaMorphometrics/Data/RapaNuiMataaData.txt" 
# We read the whole file
PAST <- read.table(data.path, header = TRUE, sep="\t")
last.meta <- which(colnames(PAST)=="X1") - 1
fac <- PAST[, 1:last.meta]
# xy will contain coordinates only
xy <- as.matrix(PAST[, -c(1:last.meta)])
# a short loop to reorder thing and store them in a list
coo <- list()
for (i in 1:nrow(xy)){ 
  coo[[i]] <- cbind(xy[i, seq(1, ncol(xy), 2)], xy[i, seq(2, ncol(xy), 2)])}
# we renames the components of the list (ie the shapes)
names(coo) <- fac[, 1]

numberOfMataa <- nrow(xy)
#maximum width
xMaxLength<-0
for (i in 1:nrow(xy)){ xMaxLength[i]<-(max(xy[i, seq(2, ncol(xy), 2)])-min(xy[i, seq(2, ncol(xy), 2)]) ) }

#maximum width
xMaxWidth<-0
for (i in 1:nrow(xy)){ xMaxWidth[i]<-(max(xy[i, seq(1, ncol(xy), 2)])-min(xy[i, seq(1, ncol(xy), 2)]) ) }
```

In order to test these hypotheses, we used morphometric outline analysis. With this technique exact outlines of artifacts can be overlaid and compared making variations between artifacts readily apparent. Our assemblage consisted of planview photographs of (N='r numberOfMataa') *mata’a* from 2 museum collections. The first museum collection consisted of XXX *mata'a* housed at the P. Sebastian Englert Museum on Rapa Nui.  This collection is composed of XX *mata'a* collected from XX locations on the island and XXX *mata'a* for which provience is known only to the level of the island itself. [[More detail on the locations?]]

The second collection of *mata'a* is composed of XXX objects housed at the Bishop Museum, Honolulu, Hawai'i. 

of t of XXX mata'a These *mata'a* wer taken by Lipo on Rapa Nui from 2007 through 2009.  25 to 33 artifacts were randomly selected from each of (4) sites. 

Using TPSdig software (Rohlf 2014), each mata’a was measured using 200 sets of X-Y coordinates from the perimeter of the artifact. Radial measurements were also taken every 2 degrees around the circumference. 

We transferred the data to PAST 3.01 (Hammer et al. 2001)  software and X-Y coordinate and radii analyses were done. The X-Y coordinates were used to create overlying images showing shape variability. The radii analysis were used to compute the standard deviation of each radius. The standard deviations were then used to graph out areas of constraint and/or variability in each group.

Each set of analysis were performed separately for each site, and collectively for the entire assemblage.  


```{r}
# now we create the Out object (formerly Coo,
# but Coo became a super-class now, to handle outlines, open outlines, and
# landmarks)
RapaNui <- Out(coo, fac=fac)

panel(RapaNui, fac = "Site")
panel(RapaNui, fac = "Island")
panel(RapaNui, fac="Source", reorder="Source")

# Lets begin by centering, scaling them 
stack(RapaNui)

RapaF <- eFourier(RapaNui, 12)
RapaP <- fgProcrustes(RapaNui)
RapaFP <- eFourier(RapaP, 12, norm=FALSE)

#by source
plot(PCA(RapaF), "Source") # regular EFT with normalized coefficients
plot(PCA(RapaFP), "Source") # Procrustes aligned (normalization of the outlines)

# by site
plot(PCA(RapaF), "Site") # regular EFT with normalized coefficients
plot(PCA(RapaFP), "Site") # Procrustes aligned (normalization of the outlines)


```


Some embedded plots.


```{r, echo=FALSE}
plot(PCA(RapaFP), "Site") # Procrustes aligned (normalization of the outlines)
```
